On the spatial aggregation of ecosystem condition indicators

### 2025-02-19 manuscript/manuscript.qmd
latex table in pti example

@@

::: callout-note
**Box 1 -- Related concepts**\

@@

It is common in the ECA community of practice to use the term aggregation about the process of aggregating normalised indicators into indices (i.e. thematic or vertical aggregation; @langhans2014), but in this paper we are primarily discussing spatial (i.e. horizontal) aggregation.
\\

@@

The modifiable area unit problem (MAUP) describes the sensitivity of spatially aggregated information to the choice of spatial units used for data collection or aggregation [@openshaw1984, @wong2009].
These units can be arbitrary (*modifiable*), yet different sizes and configuration of units can have large effects on the outcome of aggregation.
Administrative units are often used as the ecosystem assets in SEEA EAs, yet they are in an ecological sense almost always arbitrary, and can also be subject to change over time.
The original spatial units that data are collected at may be a pragmatic choice in much the same way.
Further still, data may not be available at the original spatial scale, but may only be downloadable in preaggregated form [@vanbeurden1999].
All these cases lead to some form of aggregation error[@rastetter1992] where the aggregated information has a non-linear relationship with the true variable value.
However, the issue of MAUP in SEEA EA deserves an in-depth analyses that we cannot provide here.
\\

@@

Generally, because of the normalisation process, indicators are more aligned with the purpose of ECA with describing ecosystem condition in normative terms.
Although variable accounts are allowed under the SEEA EA framework, ans is indeed a central feature, we believe they should be seen as interim accounts for metrics that cannot or have not yet been normalised.
We would therefore argue that when an indicator has been developed, the original variable is not presented.
And when only variable estimates are given, extra care is given to describing the spatial aggregation pathway.

@@

As we have seen, aggregation displacement causes a divergence in the type of information made available through the spatial aggregation of variables versus indicators.
In addition, unavoidable aggregation displacement (for good and bad) will also affect the indicator values and the inferences that can be drawn from these.
If not careful, aggregation displacement may also cause errors to sneak into the indicator logic.
A major determinant of the resulting aggregation displacement from the spatial aggregation of ecosystem condition indicators (@fig-agg1), is the choice about the order of the different steps in the normalisation and aggregation process, i.e. the aggregation pathway (@fig-pathways).
If normalisation occurs early in the aggregation process, the amount of compensation, or off-setting, is limited (**Example 2**).
However, sometimes complete or partial compensation is justified (**Example 1 and 3**).

@@

<!-- Dataset: “Vannmiljø”, the Norwegian Environment Agency’s database for Water Framework Directive relevant measurements -->

<!-- Reference: https://github.com/NINAnor/NI_vannf -->

*Variable*: Mean score of algal species present, based on a set of indicator species scored for phosphorus requirements/tolerance.
The variable is recorded in water bodies (lakes of 0,5 km^2^ or more).

@@

The Water Framework Directive (WFD) indicators uses aggregation pathway 7 (@fig-pathways), with truncation, scaling, and transformation, in that order.
The value obtained after the first two steps is called an EQR (ecological quality ratio), and the value obtained after the third step nEQR (normalised EQR).
Spatial aggregation, the way that we use the term here, is not done for the WFD.
Instead, the proportion of water bodies in each condition class is the aggregated reporting metric.
Therefore aggregation displacement is not an issue.
When WFD data are put into use in other contexts, however, aggregation becomes important.\

Because of how X~100~ is defined based on the median value across reference lakes, overshooting values (\> X~100~) should be preserved in the spatial aggregation.
Otherwise we get a negative displacement, making it practically impossible to reach an indicator value of 1.

This means that neither EQR values, nor nEQR values, can be uncritically aggregated spatially to be used in a SEEA EA compliant in ECA.
One solution, to enable the use of WFD indicators in ECAs, would be to change to pathway 6, with (1) scaling, (2) transforming, (3) aggregating and (4) truncating the values [@fig-pathways].\

Furthermore, Eurostat [@eurostat2022] encourages reporting of variable values.
These, however, cannot be aggregated spatially in any meaningful way for WFD indicators.
The issue is the same as for all other variables for which normalised indicators exist, but is perhaps more extreme for WFD indicators due to different reference conditions and levels for different water body types, and strong non-linear transformation functions that inflate the aggregation displacement issue.
<!-- Can we have a figure of calculation exmaple here? -->
:::

>>>>

```{r setup, include=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
```

@@

::: {.callout-note}
**Box 1 -- Related concepts**\

@@

It is common in the ECA community of practice to use the term aggregation about the process of aggregating normalised indicators into indices (i.e. thematic or vertical aggregation; @langhans2014), but in this paper we are primarily discussing spatial (i.e. horizontal) aggregation. \

@@

The modifiable area unit problem (MAUP) describes the sensitivity of spatially aggregated information to the choice of spatial units used for data collection or aggregation [@openshaw1984, @wong2009].
These units can be arbitrary (*modifiable*), yet different sizes and configuration of units can have large effects on the outcome of aggregation.
Administrative units are often used as the ecosystem assets in SEEA EAs, yet they are in an ecological sense almost always arbitrary, and can also be subject to change over time.
The original spatial units that data are collected at may be a pragmatic choice in much the same way.
Further still, data may not be available at the original spatial scale, but may only be downloadable in preaggregated form [@vanbeurden1999].
All these cases lead to some form of aggregation error[@rastetter1992] where the aggregated information has a non-linear relationship with the true variable value.
However, the issue of MAUP in SEEA EA deserves an in-depth analyses that we cannot provide here. \

@@

Generally, because of the normalisation process, indicators are more aligned with the purpose of ECA with describing ecosystem condition in normative terms.
Although variable accounts are allowed under the SEEA EA framework, and is indeed a central feature, we believe they should be seen as interim accounts for metrics that cannot or have not yet been normalised.
We would therefore argue that when an indicator has been developed, the original variable is not presented.
And when only variable estimates are given, extra care is given to describing the spatial aggregation pathway.
Furthermore, Eurostat [@eurostat2022] encourages reporting of variable values.
Sometimes, however, variable values cannot be aggregated spatially in any meaningful way (see **Example 3**).

@@

As we have seen, aggregation displacement causes a divergence in the type of information made available through the spatial aggregation of variables versus indicators.
In addition, unavoidable aggregation displacement (for good and bad) will also affect the indicator values and the inferences that can be drawn from these.
If not careful, aggregation displacement may also cause errors to sneak into the indicator logic.
A major determinant of the resulting aggregation displacement from the spatial aggregation of ecosystem condition indicators (@fig-agg1), is the choice about the order of the different steps in the normalisation and aggregation process, i.e. the aggregation pathway (@fig-pathways).
If normalisation occurs early in the aggregation process, the amount of compensation, or off-setting, is limited (**Example 2**).
However, sometimes complete or partial compensation is justified (**Example 1** and **3**).

@@

*Variable*: Mean score of algal species present, based on a set of indicator species scored for phosphorus requirements/tolerance.
The variable is recorded in water bodies (lakes of 0.5 km^2^ or more).

@@

```{r read}
#| include: false
dat <- readxl::read_xlsx(here::here("data/PTI_data.xlsx")) |>
  mutate(X = X |> str_replace_all("%", "\\\\%"))
cn <- colnames(dat) |>
    str_replace_all("2", "") |>
    str_replace_all("X", "") |>
    str_replace_all("RL", "X\\\\textsubscript{100}")
```

```{r tbl-pti}
#| echo: false
#| tbl-cap: "Dummy data for the PTI example. value = value for a hypothetical variable. agg. = spatial aggregation (arithemtic mean); trunc. truncation; stat. = spatial statistics. The transformation step is left out for simplicity. Pathways refer to figure 3."

kbl(dat, format = "latex",
    booktabs = TRUE,
    col.names = cn,
    escape = F
    ) |>
  kable_styling() %>%
  add_header_above(c(
    " " = 3, 
    "Pathway 5" = 4,
     "Pathway 6" = 3)) |>
  collapse_rows(
    columns = c(9,10),
    latex_hline = "major"
  )

The Water Framework Directive (WFD) indicators uses aggregation pathway 5 (@fig-pathways), with truncation, scaling, and transformation, in that order (see worked example in @tbl-pti).
The value obtained after the first two steps is called an EQR (ecological quality ratio), and the value obtained after the third step nEQR (normalised EQR).
Spatial aggregation, the way that we use the term here, is not done for the WFD.
Instead, the proportion of water bodies in each condition class is the aggregated reporting metric.
Therefore aggregation displacement is not an issue.
When WFD data are put into use in other contexts, however, aggregation becomes important.

\
Because of how X~100~ is defined based on the median value across reference lakes, overshooting values (\> X~100~) should be preserved in the spatial aggregation.
Otherwise we get a negative displacement, making it practically impossible to reach an indicator value of 1.
This means that neither EQR values, nor nEQR values, can be uncritically aggregated spatially to be used in a SEEA EA compliant in ECA.
One solution, to enable the use of WFD indicators in ECAs, would be to change to pathway 6, with (1) scaling, (2) transforming, (3) aggregating and (4) truncating the values.
In our hypothetical example (@tbl-pti) this gives and aggregated value of 0.85, whereas the mean value for the nEQRs would be 0.75 (i.e. negatively displaced).
:::
